# Tuesday Exercises

## Question 1

### (a)

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
calcium_data <- read.table(file = "../../data/T3_3_Calcium.DAT", sep = "")
calcium_data
names(calcium_data) <- c("location","avail","exch","turnip-green-calc")
calcium_data
```

### (b)

```{r}
cov_mat <- cov(calcium_data[,2:4])
cov_mat
```

```{r}
cor_mat <- cor(calcium_data[,2:4])
cor_mat
```

```{r}
mean_vec <- colMeans(calcium_data[,2:4])
mean_vec
```

-   Variance of the turnip green calcium level is 0.25. Standard deviation of it is 0.5. Covariance between available and exchangeable is 49.68.

-   They have a linear positive association but not sure due its magnitude being tough to interpret. A better comparison would be the correlation coefficient in the correlation matrix where the value is 0.49. This shows that there is a moderate linear positive association.

-   More strongly associated with exchangeable with a value of 0.86.

### (c)

Make a scatterplot matrix of the 3 variables.

```{r}
pairs(calcium_data[,2:4])
```

From this it would seem that between avail and exch there is somewhat weak positive linear association. Between exch and turnip green levels there is a better trend. And with turnip green and avail there is also a weak association between the variables.

### (d)

```{r}
# scatterplot3d function requires scatterplot3d library. Also add color to the points by what data point they are/
library(scatterplot3d)
scatterplot3d(x=calcium_data$avail, y=calcium_data$exch, z=calcium_data$`turnip-green-calc`, color=calcium_data$location)

```

From this 3D graph, its is somewhat hard to tell what patterns exist in the data since they are not grouped by anything meaningful other than location.

### (e)

```{r}
stars(calcium_data[,2:4], draw.segments="T", labels=calcium_data$location, key.loc=c(6, 1.8))
```

-   Location 3 has massive values for all 3 variables: exch, turnip green and avail calcium levels

-   Locations seem to have a lot of avail calcium levels.

## Question 2

### (a)

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
probes <- read.table(file="../../data/T3_5_Probe.DAT", sep="")
probes
names(probes) <- c("subject_no","y1","y2","y3","y4","y5")
probes
```

### (b)

Create a parallel coordinates plot of the 5 response times for the 11 subjects.

```{r}
library(MASS)
parcoord(probes, col=1:11, lty=1:11, main="Parallel Coordinates Plot of Response Times", var.label=TRUE)

```

There does not seem to be any groupings in the data, which means it looks like it is a random jumble of responses for each of the subjects.

### (c)

Fake grouping variable where first 45 records were for younger adults and last 6 were for older adults.

```{r}
age <- c(rep(1, 5), rep(2, 6))
```

Generate a parallel coordinates plot where younger adults are in red and older adults are in blue.

```{r}
colors <- c("Red", "Blue")
parcoord(probes, col=colors[age], lty=1:11, main="Parallel Coordinates Plot of Response Times", var.label=TRUE)
```
